Visualizing Contingency Tables

### 2006-05-03 trunk/vcd/paper/hsc.Rnw
KH improvements.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@528 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

As an example, consider the well-known UCB admissions data 
\cite{vcd:Bickel+Hammel+O'Connell:1975}
on applicants to graduate school at the University of California in Berkeley for
the six largest departments in 1973 classified by admission and
gender. The `flattened' contingency table, putting department into the
columns and admit---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all
departments gives wrong evidence of gender bias, 
i.e., higher admission rates for male students.

@@

A first step in the exploratory analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.
In particular, departments A and B have a higher proportion of male students
and a lower rejection rate compared to the other departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is use a mosaic display
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the tiles it can be seen that an
unequal number of students apply for the six departments (with a particularly small
number of females in departments A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

A first approach is to use pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
% independence in the upper triangle, 
% mosaics for \emph{joint} independence 
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately 
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level 
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression.
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by using the conditioning variables first for splitting.
However, this also displays the marginal distribution of the conditioning variables
which might make it more difficult to interpret the dependent variables of
interest. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to complement the conditioning in the model by
a conditioning in the plot, i.e., to use trellis-like layouts for
visualizing \emph{partial} tables as defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding plots in the panels have the same size. 
In addition, trellis layouts help to reduce the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. In Figure~\ref{fig:ucbadmissions},
data-driven cut-offs (derived from the maximum test for conditional independence)
are used, resulting in a more colorful shading than the fixed cut-offs in
Figure~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}.

@@

In summary, both age and education clearly matter here.
But is there really any significant influence of memory on the attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
Are memory and attitude conditionally independent, given age and education?
To focus on this question, we will take another model-based view on the data
that avoids the distracting influence of the marginal variables (age
and education): We employ
partial mosaic plots in a trellis layout (conditioning on age and education) with
residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
This visualization brings out much more clearly that the increased approval rate
among people with memories of punishments is present mainly in the elementary education
column and, to a lesser degree, in the age 40+ row. Using, e.g., a 
permutation test similar to the one discussed in Section \ref{sec:test}, 
model statistics reveal that the hypothesis is rejected at a significance level of 
1\%, but the shading clearly shows that this association is
only significant in two cells (the two older age groups with elementary education)
where fewer people without memories do not approve punishment than expected under
(conditional) independence. In the group of people with age between
25--39, the lighter shading indicates that here,
the association is only significant at the 10\% level.

@@

This chapter reviews several alternatives for the visualization of
multi-way contingency tables. For two-way tables, mosaic, sieve, and
association plots are suitable for the visualization of observed and
expected values and the Pearson residuals, respectively. This basic
methods can be enhanced by using residual-based shadings, preferably
based on perceptual color palettes such as those derived from the HCL
space. Residual-based shadings can be used to visualize sign and size
of the residuals, as well as the significance of test statistics such
as the $\chi^2$ or the maximum test statistic. The latter has the
advantage of detecting residuals causing the rejection of the hypothesis
of independence. The methods directly extend to the multi-way case by
using `flat' representations of the multi-way tables, and specialized
displays for conditional independence such as trellis layouts of
partial tables and pairs plots.

>>>>

As an example, consider the well-known UCB admissions data
\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
admission and gender, to graduate school at the University of California
in Berkeley for the six largest departments in 1973. The `flattened'
contingency table, putting department into the columns and
admit---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all departments gives
wrong evidence of gender bias, i.e., higher admission rates for male
students.

@@

A first step in the exploratory analysis of more complex tables is to
get a quick overview on the data. For this, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells feature plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.
In particular, departments A and B have a higher proportion of male students
and a lower rejection rate compared to the other departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is to use a mosaic
display, 
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the
tiles it can be seen that students apply to the six departments in
unequal numbers (with a particularly small
number of females in departments~A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

One approach is employing pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
% independence in the upper triangle, 
% mosaics for \emph{joint} independence 
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately 
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level 
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression.
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to obtain a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by  first using the conditioning variables for splitting.
However, this also displays the marginal distributions of the conditioning variables
which might make it more difficult to interpret the dependent variables of
interest. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to complement the conditioning in the model by
a conditioning in the plot, i.e., to use trellis-like layouts for
visualizing \emph{partial} tables as defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding plots in the panels have the same size. 
In addition, trellis layouts help to reduce the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. In Figure~\ref{fig:ucbadmissions},
data-driven cut-offs (derived from the maximum test for conditional independence)
are used, resulting in a more colorful shading than the fixed cut-offs in
Figure~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}.

@@

In summary, both age and education clearly matter here.
But is there really any significant influence of memory on the attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
Are memory and attitude conditionally independent, given age and education?
To focus on this question, we take another model-based view on the data
which avoids the distracting influence of the marginal variables (age
and education): We employ
partial mosaic plots in a trellis layout (conditioning on age and education) with
residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
This visualization brings out much more clearly that the increased approval rate
among people with memories of punishments is present mainly in the elementary education
column and, to a lesser degree, in the age 40+ row. Using, e.g., a 
permutation test similar to the one discussed in Section \ref{sec:test}, 
model statistics reveal that the hypothesis is rejected at a significance level of 
1\%, but the shading clearly shows that this association is
only significant in two cells (the two older age groups with elementary education)
where fewer people without memories do not approve punishment than expected under
(conditional) independence. In the group of people with age between
25--39, the lighter shading indicates that here,
the association is only significant at the 10\% level.

@@

This chapter reviews several alternatives for the visualization of
multi-way contingency tables. For two-way tables, mosaic, sieve, and
association plots are suitable for the visualization of observed and
expected values and the Pearson residuals, respectively. These basic
methods can be enhanced by using residual-based shadings, preferably
based on perceptual color palettes such as those derived from the HCL
space. Residual-based shadings can be used to visualize sign and size
of the residuals, as well as the significance of test statistics such
as the $\chi^2$ or the maximum test statistic. The latter has the
advantage of detecting residuals causing the rejection of the hypothesis
of independence. The methods directly extend to the multi-way case by
using `flat' representations of the multi-way tables, and specialized
displays for conditional independence such as trellis layouts of
partial tables and pairs plots.
